---
title: "Simulated Gamma distribution example"
output: 
  rmarkdown::html_vignette:
       toc: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Simulated Gamma distribution example}
  %\usepackage[UTF-8]{inputenc}
---

## Overview

```{r, setup, message = FALSE, warning = FALSE}

library(tidyverse)
library(boundHTS)
library(forecast)
```


This vignette demonstrates how to:

1. Simulate correlated Gamma distributed time series at the bottom level of a hierarchy
2. Aggregate these series to higher levels with additional variability
3. Estimate Bayesian state–space models at each level using auto-arima.
4. Perform rolling one-step-ahead forecasting efficiently by reusing compiled models

The example is intentionally simple (two bottom-level series and one aggregate), but the workflow generalizes to larger hierarchies.

### Hierarchical structure

We consider a two-level hierarchy:

- **Bottom level**: two correlated proportions (`AA`, `AB`)
- **Top level**: their aggregate (`Total`).

All series take values in (0, Inf) and are modelled using Gamma distributions. 


## Step 1: Load simulated Beta distributed data 

```{r}
head(gamma_sim_data)

```


## Step 2: Generating model estimates

We estimate separate state–space models for:

- the **top-level series** (`Total`)
- the **mid-level series** (`A`, `B`)
- the **bottom-level series** (`AA`, `AB`, `BA`, `BB`)

### 2.1 Run the models

```{r}
nsim <- 1000
n_testyears <- nrow(gamma_sim_data) # withhold one year of data as an example

# Split data into test and training
train_idx <- which(gamma_sim_data$Time < n_testyears)
training_data <- gamma_sim_data[train_idx, , drop = FALSE]
  
# Fit models
mod_A_logit <- forecast::auto.arima(training_data$logit_A) 
mod_AA_logit <- forecast::auto.arima(training_data$logit_AA) 
mod_AB_logit <- forecast::auto.arima(training_data$logit_AB) 

# Get one-step ahead predictions
fc_A <- as_tibble(forecast::forecast(mod_A_logit, h = 1))
fc_AA <- as_tibble(forecast::forecast(mod_AA_logit, h = 1))
fc_AB <- as_tibble(forecast::forecast(mod_AB_logit, h = 1))
  
# Get variance of one-step ahead forecast
sim_logit_A <- replicate(5000, simulate(mod_A_logit, nsim = 1, future = TRUE))
var_beta_A  <- var(sim_logit_A)
sim_logit_AA <- replicate(5000, simulate(mod_AA_logit, nsim = 1, future = TRUE))
var_beta_AA  <- var(sim_logit_AA)
sim_logit_AB <-  replicate(5000, simulate(mod_AB_logit, nsim = 1, future = TRUE))
var_beta_AB  <- var(sim_logit_AB)
  
# Create results dataframes for point estimates and variance
fc_A <- fc_A %>% mutate(Node = "A",
                        Time = n_testyears,
                        logit_var = var_beta_A) %>%
  rename(logit_mean = `Point Forecast`)
fc_AA <- fc_AA %>% mutate(Node = "AA",
                          Time = n_testyears,
                          logit_var = var_beta_AA)  %>%
    rename(logit_mean = `Point Forecast`)
fc_AB <- fc_AB %>% mutate(Node = "AB",
                          Time = n_testyears,
                          logit_var = var_beta_AB)  %>%
  rename(logit_mean = `Point Forecast`)

forecasts <- bind_rows(fc_A, fc_AA, fc_AB) %>%
  mutate(Node = factor(Node, levels = c("AA", "AB", "A"))) %>%
  arrange(Node)
 

```


### 2.3 Back-transform the proportions
```{r}
# Back transform to proportions
prop_var <- inv_logit_delta(inv_logit(forecasts$logit_mean), forecasts$logit_var)

forecasts <- forecasts %>%
  mutate(proportion = inv_logit(forecasts$logit_mean),
         prop_var = prop_var) %>%
  select(Time, Node, logit_mean, logit_var, proportion, prop_var)

```


## Step 3: Coherent hierarchical forecasting

We estimate a coherent top-series probability density function (PDF) using a convolution and weighted Beta distributions. As inputs into this process we use:
- the **top-level series** (`A`) estimates
- the **bottom-level series** (`AA`, `AB`) estimates
- A fine grid over which the density exists
- Bottom-series weights

### Step 3.1: Generate probability density function via convolution
We will generate PDFs for the withheld data points to validate our method. You can loop over all years of you wish by updating the code to do so.

```{r}
## Get Beta parameters
beta_params_df <- get_beta_params(forecasts$proportion, forecasts$prop_var)
beta_params_df$Node <- forecasts$Node
  
alpha_list <- list(beta_params_df %>% filter(Node %in% c("AA", "AB")) %>% pull(Alpha),
                   beta_params_df %>% filter(Node %in% c("A")) %>% pull(Alpha))

beta_list <- list(beta_params_df %>% filter(Node %in% c("AA", "AB")) %>% pull(Beta),
                   beta_params_df %>% filter(Node %in% c("A")) %>% pull(Beta))

# Set up weights
weights_list <- list(c(0.75, 0.25), 1)

# Set up expected values 
mu_theory <- forecasts %>% 
  filter(Node %in% c("A")) %>% 
  arrange(Node) %>% 
  pull(proportion)

# Run convolution with tilting 
res <- run_convolution_with_tilting(family = "Beta",
                                groups =  list(Bottom = c("AA", "AB"), Top = "A"),
                                point = TRUE,
                                mu_theory = mu_theory,
                                z_values = seq(0, 1, length.out=1000),
                                alpha_list = alpha_list,
                                beta_list = beta_list,
                                weights_list = weights_list,
                                n_draws = 500,
                                n_sims = 50)

```


## Step 4: Visualise the tilted convoluted PDF and compare against the predictive mean

The figure presents the predictive density formed by convolving the estimated bottom-level weighted Beta distributions. The resulting density is subsequently tilted so that its expectation matches the corresponding one-step-ahead predictive mean, ensuring coherence between the full predictive distribution and the point forecast.

For a single time point:
```{r, echo=FALSE, fig.width=9, fig.height=5, dpi=150}
t_point <- 250

obs_point <- beta_sim_data %>% 
  filter(Time==t_point) %>%
  mutate(Method="Observed data")

forecasts_mean <- forecasts %>% 
  dplyr::filter(Time == t_point) %>%
  dplyr::select(Time, Node, proportion) %>%
  tidyr::pivot_wider(names_from = Node, values_from = proportion) %>%
  dplyr::mutate(Method = 'Predicted mean')

exp_plot_df <- res$tilted_density[[1]] %>% mutate(Method= "Tilted density")


ggplot2::ggplot() +
  ggplot2::geom_line(data = exp_plot_df,
                ggplot2::aes(x = Z, y = Density, colour=Method), linewidth = 1) +
  ggplot2::geom_vline(data = forecasts_mean,
                ggplot2::aes(xintercept = A, colour = Method), linewidth = 1) +
  ggplot2::geom_vline(data = obs_point,
                ggplot2::aes(xintercept = A, colour = Method), linewidth = 1) +
  ggplot2::labs(x = "Proportion", y = "Density") +
  ggplot2::theme_minimal(base_size = 14) +
  ggplot2::theme(legend.position = "bottom")

```

