---
title: "Simulated Poisson distribution example"
output: 
  rmarkdown::html_vignette:
       toc: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Simulated Poisson distribution example}
  %\usepackage[UTF-8]{inputenc}
---

## Overview

```{r, setup, message = FALSE, warning = FALSE}
set.seed(123)
```


This vignette demonstrates how to:

1. Forecast simulated Poisson-distributed time series using `glm` from the \CRANpkg{stats}.
2. Perform rolling one-step-ahead forecasting efficiently by reusing compiled models

The example is intentionally simple (two bottom-level series and one aggregate), but the workflow generalizes to larger hierarchies.

### Hierarchical structure
In this example, we use a simple simulated hierarchical count dataset with four bottom-level series.  Each series is generated from a Poisson GLM with quadratic dependence on a covariate `x`. 

We consider a two-level hierarchy:

- **Bottom level**: four correlated counts (`AA`, `AB`, `BA`, `BB`)
- **Top level**: their aggregate (`Total`) with noise.


## Step 1: Load and visualise Poisson distributed data 

```{r}
sim_data <- boundHTS::poisson_sim_data

```

This shows the counts recorded over the covariate X for each node in the hierarchy.
```{r}
hts_data_long <- tidyr::pivot_longer(sim_data, cols = -c(X), names_to = "Level", values_to = "Value")

ggplot2::ggplot(hts_data_long, ggplot2::aes(x = X, y = Value, color = Level)) +
  ggplot2::geom_line() +
  ggplot2::labs(title = "Simulated hierarchical count", x = "X", y = "Count") +
  ggplot2::theme_minimal() + 
  ggplot2::theme(legend.position = "none") +
  ggplot2::facet_wrap(~Level, scales='free')

```


## Step 2: Generating model estimates

We now fit **Poisson GLMs** to each bottom-level count series (`AA`, `AB`, `BA`, `BB`). Each model uses a quadratic function of the covariate `X` and is estimated in a **rolling one-step-ahead forecasting framework**.

At each forecast origin:

1. The model is trained using all observations up to time `t − 1`
2. The Poisson rate (`lambda`) is estimated in-sample
3. A one-step-ahead forecast of `lambda` is produced
4. Predictive samples are drawn from a Poisson distribution

### Step 2.1 Set up modelling variables 
```{r}
N <- nrow(sim_data) # 2000 rows
m <- 4 # bottom series
n_series <- ncol(sim_data[,-1])
n_train <- c(1:c(N-50)) # withhold the last 50 observations for validation
test_indices <- c(length(n_train)+1):N
n_samples <- length(test_indices)
sum_bottom <- c("AA", "AB", "BA", "BB")
top_y_vals <- seq(from = 0, to = c(max(sim_data$Tot)+10))
ally <- sim_data[,-c(1)] # remove covariates

```


### Step 2.2 Rolling Poisson GLM estimation

```{r, poisson-regression}
# Containers for results
poiss_reg <- list()   # fitted GLMs at each step
poiss_GLM_reg <- list()
lambda_list <- list()   # estimated Poisson rates (lambda)
fitted_list <- list()   # predictive Poisson samples

# Bottom-level series
sum_bottom <- c("AA", "AB", "BA", "BB")

for (t in seq_along(test_indices)) {

  # Training indices (up to time t-1)
  train_idx <- seq_len(test_indices[t] - 1)

  # Store estimated lambdas and predictive samples
  lambda_est <- fits <- matrix(
    NA,
    nrow = length(train_idx) + 1,
    ncol = n_series
  )

  colnames(lambda_est) <- colnames(fits) <- colnames(ally)

  # Fit Poisson GLM separately for each bottom-level series
  for (series in colnames(ally)) {

    # Quadratic Poisson regression
    formula_str <- paste(series, "~ X + I(X^2)")
    fit <- stats::glm(
      formula = stats::as.formula(formula_str),
      data    = sim_data[train_idx, c("X", series)],
      family = "poisson"
    )

    # Store fitted model
    poiss_reg[[series]] <- fit

    # In-sample fitted Poisson means
    lambda_est[train_idx, series] <- fit$fitted.values

    # One-step-ahead Poisson mean
    lambda_est[test_indices[t], series] <-
      stats::predict(fit, newdata = sim_data[test_indices[t], c("X", series)])

    # Draw predictive samples from Poisson distribution
    fits[, series] <- stats::rpois(
      n = test_indices[t],
      lambda = lambda_est[, series]
    )
  }

  # Save results for this forecast origin
  lambda_list[[t]] <- lambda_est
  fitted_list[[t]] <- fits
  poiss_GLM_reg[[t]] <- poiss_reg
}

```

At the end of this procedure:

- `poiss_GLM_reg`: contains the fitted Poisson GLMs at forecast time t.
- `lambda_list`: contains estimated Poisson means (lambda) for each series, including the one-step-ahead value.
- `fitted_list`: contains simulated predictive draws from the Poisson distribution.


## Step 3: Coherent hierarchical forecasting

We estimate a coherent top-series probability mass function (PMF) using a convolution and Poisson distributions. As inputs into this process we use:
- the **top-level series** (`A`) estimates
- the **bottom-level series** (`AA`, `AB`, `BA`, `BB`) estimates
- A fine grid over which the density exists

#### Step 3.1: Set up inputs for convolution
We will generate PMFs for the withheld data points to validate of our method. You can loop over all years of you wish by updating the code to do so.

```{r, eval=FALSE}
f_tilde_exp <- list()
nu_exp <- list()
f_y <- vector()
bottom_sum <- c("AA", "AB", "BA", "BB")
pmf_values <- vector()
```


#### Step 3.2: Generate PMF via convolution and tilt the convoluted PMF towards the predictive mean.
```{r, eval=FALSE}
# Apply to predictions
for(t in 1:length(test_indices)) {
  # lambda values
  lambda_vals <- as.data.frame(lambda_list[[t]])
  mu_theory <- as.vector(unlist(lambda_vals[test_indices[t], ])) # predictive mean
  
  lambda_bseries <- lambda_vals[test_indices[t],bottom_sum]
  lambda_conv <- sum(lambda_bseries) # sum poissons
  lambda_vec <- c(lambda_conv, as.numeric(lambda_bseries)) # convolution and bottom series lambda
  
  # fitted values
  fitted_vals <- tibble::as_tibble(fitted_list[[t]])
  fitted_bseries <- fitted_vals[test_indices[t],bottom_sum]
  
  # Construct tilted pmf for top and bottom series
  f_tilt <- matrix(NA, nrow = length(top_y_vals), ncol = length(lambda_vec))
  for(k in 1:length(lambda_vec)) {
    # Convolution step
    f_y <- stats::dpois(top_y_vals, lambda_vec[k]) # density of convolution
    
    # Solve for nu
    nu_star <- stats::uniroot(moment_condition_tilting, interval = c(-1, 1), 
                       f_y = f_y, y_vals = top_y_vals, 
                       mu_theory = mu_theory[k])$root

    f_tilt[,k] <- tilted_density_discrete(nu_star, f_y, top_y_vals)
  }
  colnames(f_tilt) <- colnames(fitted_vals)
  
  nu_exp[[t]] <- nu_star
  f_tilde_exp[[t]] <- f_tilt # density of top level 
}

```


#### Step 3.3: Visualise the tilted convoluted PDF and compare against the predictive mean

The figure presents the predictive density formed by convolving the estimated bottom-level Poisson distributions. The resulting density is subsequently tilted so that its expectation matches the corresponding one-step-ahead predictive mean, ensuring coherence between the full predictive distribution and the point forecast.

```{r,  echo=FALSE, out.width = "100%",  fig.align = "center"}

t=50

t_idx <- test_indices[t]

# Extract predictive means at time t
lambda_mat <- lambda_list[[t]]
lambda_t   <- lambda_mat[t_idx, ]

lambda_bottom <- lambda_t[-1]
  
# density of convolution 
conv_top_dens <- dpois(top_y_vals, sum(lambda_bottom))

conv_plot <- data.frame(
   y = top_y_vals,
   pmf = conv_top_dens,
   node = "Tot"
)

# tilted density
df_plot <- data.frame(
  y = rep(top_y_vals, ncol(f_tilde_exp[[t]])),
  pmf = as.vector(f_tilde_exp[[t]]),
  node = rep(colnames(f_tilde_exp[[t]]), each = length(top_y_vals))
)

means_df <- data.frame(
  node = names(lambda_t),
  mean = lambda_t
  ) 
  

ggplot() +
  geom_line(data = df_plot, aes(x = y, y = pmf, col='Tilted density')) +
  geom_line(data = conv_plot, aes(x = y, y = pmf, col='Convoluted denisty')) +
  geom_vline(data = means_df,
             aes(xintercept = mean, linetype = "Predicitive mean")) +
  facet_wrap(~ node, nrow=5) +
  scale_colour_brewer(palette="Set1") +
  labs(x = "Counts",
       y = "Tilted probability mass")

```




