---
title: "Simulated Poisson distribution example"
output: 
  rmarkdown::html_vignette:
       toc: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Simulated Poisson distribution example}
  %\usepackage[UTF-8]{inputenc}
---

## Overview

This vignette demonstrates how to:

1. Forecast simulated Poisson-distributed time series using `glm` from the \CRANpkg{stats}.
2. Perform rolling one-step-ahead forecasting efficiently by reusing compiled models

The example is intentionally simple (two bottom-level series and one aggregate), but the workflow generalizes to larger hierarchies.

### Hierarchical structure
In this example, we use a simple simulated hierarchical count dataset with four bottom-level series.  Each series is generated from a Poisson GLM with quadratic dependence on a covariate `x`. 

We consider a two-level hierarchy:

- **Bottom level**: four correlated counts (`AA`, `AB`, `BA`, `BB`)
- **Top level**: their aggregate (`Total`) with noise.


## Step 1: Load and visualise Poisson distributed data 

```{r, echo=TRUE, results='hide'}
set.seed(123)
library(boundHTS)
library(tidyverse)

data("poisson_sim_data")         # load data
head(poisson_sim_data)

```

This shows the counts recorded over time for each node in the hierarchy.
```{r}
hts_data_long <- tidyr::pivot_longer(poisson_sim_data, cols = -c(Time), names_to = "Level", values_to = "Value")

ggplot2::ggplot(hts_data_long, ggplot2::aes(x = Time, y = Value, color = Level)) +
  ggplot2::geom_line() +
  ggplot2::labs(title = "Simulated hierarchical count", x = "Time", y = "Count") +
  ggplot2::theme_minimal() + 
  ggplot2::theme(legend.position = "none") +
  ggplot2::facet_wrap(~Level, scales='free')

```


## Step 2: Generating model estimates

We now fit **Poisson GLMs** to each bottom-level count series (`AA`, `AB`, `BA`, `BB`). Each model uses an AR(1) model and is estimated in a **rolling one-step-ahead forecasting framework**.

At each forecast origin:

1. The model is trained using all observations up to time `t − 1`
2. The Poisson rate (`lambda`) is estimated in-sample
3. A one-step-ahead forecast of `lambda` is produced
4. Predictive samples are drawn from a Poisson distribution

### Step 2.1 Set up modelling variables 
```{r}
N <- nrow(poisson_sim_data) # 2000 rows
m <- 4 # bottom series
n_series <- ncol(poisson_sim_data[,-1])
n_train <- c(1:c(N-50)) # withhold the last 50 observations for validation
test_indices <- c(length(n_train)+1):N
n_samples <- length(test_indices)
nodes <- c("Tot", "AA", "AB", "BA", "BB")
sum_bottom <- c("AA", "AB", "BA", "BB")
top_y_vals <- seq(from = 0, to = c(max(poisson_sim_data$Tot)+10))
ally <- poisson_sim_data[,-c(1)] # remove covariates

```


### Step 2.2 Rolling Poisson GLM estimation

```{r, poisson-regression}
# Containers for results
lambda_list <- list()
fitted_list <- list()
poiss_GLM_reg <- list()

for(t in seq_along(test_indices)) {

  my_train <- seq_len(test_indices[t]-1)

  lambda_est <- matrix(NA, nrow = length(my_train) + 1, ncol = length(nodes))
  fits <- matrix(NA, nrow = length(my_train) + 1, ncol = length(nodes))
  colnames(lambda_est) <- colnames(fits) <- nodes

  poiss_reg <- list()

  for(i in nodes) {

    y_train <- poisson_sim_data[my_train, ] %>% pull(all_of(i))

    # Fit tsglm on training series
    fit <- tscount::tsglm(y_train, model = list(past_obs = 1), link = "log")
    poiss_reg[[i]] <- fit

    # Fitted values for the training period
    lambda_est[1:length(my_train), i] <- fitted(fit)

    # One-step-ahead forecast
    lambda_est[length(my_train)+1, i] <- predict(fit, n.ahead = 1)$pred

    # Simulate Poisson counts based on lambda
    fits[, i] <- rpois(length(lambda_est[, i]), lambda = lambda_est[, i])
  }

  lambda_list[[t]] <- lambda_est
  fitted_list[[t]] <- fits
  poiss_GLM_reg[[t]] <- poiss_reg
}


```

At the end of this procedure:

- `poiss_GLM_reg`: contains the fitted Poisson GLMs at forecast time t.
- `lambda_list`: contains estimated Poisson means (lambda) for each series, including the one-step-ahead value.
- `fitted_list`: contains simulated predictive draws from the Poisson distribution.


## Step 3: Coherent hierarchical forecasting

We estimate a coherent top-series probability mass function (PMF) using a convolution and Poisson distributions. As inputs into this process we use:
- the **top-level series** (`A`) estimates
- the **bottom-level series** (`AA`, `AB`, `BA`, `BB`) estimates
- A fine grid over which the density exists

#### Step 3.1: Set up inputs for convolution
We will generate PMFs for the withheld data points to validate of our method. You can loop over all years of you wish by updating the code to do so.

```{r}
f_tilde_exp <- list()
nu_exp <- list()
f_y <- vector()
pmf_values <- vector()
```


#### Step 3.2: Generate PMF via convolution and tilt the convoluted PMF towards the predictive mean.
```{r}
# Apply to predictions
for(t in 1:length(test_indices)) {
  # lambda values
  lambda_vals <- as.data.frame(lambda_list[[t]])
  mu_theory <- as.vector(unlist(lambda_vals[test_indices[t], ])) # predictive mean
  
  lambda_bseries <- lambda_vals[test_indices[t],sum_bottom]
  
  # fitted values
  fitted_vals <- tibble::as_tibble(fitted_list[[t]])
  fitted_bseries <- fitted_vals[test_indices[t],sum_bottom]
  
  # Construct tilted pmf for top and bottom series
  f_y_bottom <- matrix(NA, nrow = length(top_y_vals), ncol = length(lambda_bseries))
  f_tilt <- matrix(NA, nrow = length(top_y_vals), ncol = length(mu_theory))
  nu_star <- vector()
  
  # Construct densities
  for(k in 1:length(lambda_bseries)) {
    # Bottom series
   f_y_bottom[,k] <- sapply(top_y_vals, function(x) {
      mean(stats::dpois(x = x, lambda = as.numeric(lambda_bseries[k])))})
   }
  f_y_top <- Poisson_convolution(z_values = top_y_vals,
                                 lambda_input = as.numeric(lambda_bseries),
                                 point=TRUE)
  f_y <- cbind(f_y_top, f_y_bottom)
  
  # Tilt density
  for(k in 1:length(mu_theory)) {
    tilted_dens <- tilt_density(mu_theory[k], top_y_vals, f_y[,k], discrete=TRUE) # tilt density
    f_tilt[,k] <- tilted_dens$f_tilted
  }
  nu_star <- c(tilted_dens$nu_star, nu_star)
  colnames(f_tilt) <- colnames(fitted_vals)
  
  nu_exp[[t]] <- nu_star
  f_tilde_exp[[t]] <- f_tilt # density of top level 
}

```


#### Step 3.3: Visualise the tilted convoluted PDF and compare against the predictive mean

The figure presents the predictive density formed by convolving the estimated bottom-level Poisson distributions. The resulting density is subsequently tilted so that its expectation matches the corresponding one-step-ahead predictive mean, ensuring coherence between the full predictive distribution and the point forecast.

```{r}

t=50

t_idx <- test_indices[t]

# Extract predictive means at time t
lambda_mat <- lambda_list[[t]]
lambda_t   <- lambda_mat[t_idx, ]

lambda_bottom <- lambda_t[-1]
  
# density of convolution 
conv_top_dens <- dpois(top_y_vals, sum(lambda_bottom))

conv_plot <- data.frame(
   y = top_y_vals,
   pmf = conv_top_dens,
   node = "Tot"
)

# tilted density
df_plot <- data.frame(
  y = rep(top_y_vals, ncol(f_tilde_exp[[t]])),
  pmf = as.vector(f_tilde_exp[[t]]),
  node = rep(colnames(f_tilde_exp[[t]]), each = length(top_y_vals))
)

means_df <- data.frame(
  node = names(lambda_t),
  mean = lambda_t
  ) 
  

ggplot2::ggplot() +
  ggplot2::geom_line(data = df_plot, ggplot2::aes(x = y, y = pmf, col='Tilted density')) +
  ggplot2::geom_line(data = conv_plot, ggplot2::aes(x = y, y = pmf, col='Convoluted denisty')) +
  ggplot2::geom_vline(data = means_df,
             ggplot2::aes(xintercept = mean, linetype = "Predicitive mean")) +
  ggplot2::facet_wrap(~ node, nrow=5) +
  ggplot2::scale_colour_brewer(palette="Set1") +
  ggplot2::labs(x = "Counts",
       y = "Tilted probability mass")

```




