---
title: "Simulated Poisson distribution example"
output: 
  rmarkdown::html_vignette:
       toc: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Simulated Poisson distribution example}
  %\usepackage[UTF-8]{inputenc}
---

## Overview

This vignette demonstrates how to:

1. Forecast simulated Poisson-distributed time series using `glm` from the \CRANpkg{stats}.
2. Perform rolling one-step-ahead forecasting efficiently by reusing compiled models

The example is intentionally simple (two bottom-level series and one aggregate), but the workflow generalizes to larger hierarchies.

### Hierarchical structure
In this example, we use a simple simulated hierarchical count dataset with four bottom-level series.  Each series is generated from a Poisson GLM with quadratic dependence on a covariate `x`. 

We consider a two-level hierarchy:

- **Bottom level**: four correlated counts (`AA`, `AB`, `BA`, `BB`)
- **Top level**: their aggregate (`Total`) with noise.


## Step 1: Load and visualise Poisson distributed data 

```{r, echo=TRUE, results='hide'}
set.seed(123)
library(boundHTS)
library(tidyverse)

data("poisson_sim_data")         # load data
head(poisson_sim_data)

```

This shows the counts recorded over time for each node in the hierarchy.
```{r}
hts_data_long <- tidyr::pivot_longer(poisson_sim_data, cols = -c(Time), names_to = "Level", values_to = "Value")

ggplot2::ggplot(hts_data_long, ggplot2::aes(x = Time, y = Value, color = Level)) +
  ggplot2::geom_line() +
  ggplot2::labs(title = "Simulated hierarchical count", x = "Time", y = "Count") +
  ggplot2::theme_minimal() + 
  ggplot2::theme(legend.position = "none") +
  ggplot2::facet_wrap(~Level, scales='free')

```


## Step 2: Generating model estimates

We now fit **Poisson GLMs** to each bottom-level count series (`AA`, `AB`, `BA`, `BB`). Each model uses an AR(1) model and is estimated in a **rolling one-step-ahead forecasting framework**.

At each forecast origin:

1. The model is trained using all observations up to time `t − 1`
2. The Poisson rate (`lambda`) is estimated in-sample
3. A one-step-ahead forecast of `lambda` is produced
4. Predictive samples are drawn from a Poisson distribution

### Step 2.1 Set up modelling variables 
```{r}
N <- nrow(poisson_sim_data) # 2000 rows
nodes <- c("AA", "AB", "BA", "BB", "Tot")
```


### Step 2.2 Rolling Poisson GLM estimation

```{r, poisson-regression}
# Containers for results
my_train <- N-1
lambda_est <- matrix(NA, nrow = my_train + 1, ncol = length(nodes))
fits <- matrix(NA, nrow = my_train + 1, ncol = length(nodes))
colnames(lambda_est) <- colnames(fits) <- nodes

for(i in nodes) {
  y_train <- poisson_sim_data[1:my_train, ] %>% pull(all_of(i))

  # Fit tsglm on training series
  fit <- tscount::tsglm(y_train, model = list(past_obs = 1), link = "log")

  # Fitted values for the training period
  lambda_est[1:my_train, i] <- fitted(fit)

  # One-step-ahead forecast
  lambda_est[my_train+1, i] <- predict(fit, n.ahead = 1)$pred

  # Simulate Poisson counts based on lambda
  fits[, i] <- rpois(length(lambda_est[, i]), lambda = lambda_est[, i])
}


```

At the end of this procedure:

- `fit`: contains the fitted Poisson GLMs at forecast time t.
- `lambda_est`: contains estimated Poisson means (lambda) for each series, including the one-step-ahead value.
- `fits`: contains simulated predictive draws from the Poisson distribution.


## Step 3: Coherent hierarchical forecasting

We estimate a coherent top-series probability mass function (PMF) using a convolution and Poisson distributions. As inputs into this process we use:
- the **top-level series** (`Tot`) estimates
- the **bottom-level series** (`AA`, `AB`, `BA`, `BB`) estimates
- A fine grid over which the density exists
- A summation matrix S

#### Step 3.1: Generate PMF via convolution and tilt the convoluted PMF towards the predictive mean.
```{r}
# Set up summation matrix
S <- rbind(Tot = c(1, 1, 1, 1),
           AA = c(1, 0, 0, 0),
           AB = c(0, 1, 0, 0),
           BA = c(0, 0, 1, 0),
           BB = c(0, 0, 0, 1))
colnames(S) <- c("AA", "AB", "BA", "BB")
 

# Set up expected values 
mu_theory <- unlist(lambda_est[N,]) # predictive mean of the aggregated series

# Run convolution with tilting 
res <- run_convolution_with_tilting(family = "Poisson",
                                    S = S,
                                    point = TRUE,
                                    mu_theory = mu_theory,
                                    z_values = seq(0,45),
                                    lambda_mat = lambda_est[N,],
                                    n_draws = 2000,
                                    n_sims = 100)

```


#### Step 3.3: Visualise the tilted convoluted PDF and compare against the predictive mean

The figure presents the predictive density formed by convolving the estimated bottom-level Poisson distributions. The resulting density is subsequently tilted so that its expectation matches the corresponding one-step-ahead predictive mean, ensuring coherence between the full predictive distribution and the point forecast.

```{r}
# Observed data 
obs_t <- poisson_sim_data[N, "Tot"]
obs_df <- data.frame(node = names(obs_t), observed = obs_t) 

# Extract predictive means at time t
lambda_t <- lambda_est[N, "Tot"]
means_df <- data.frame(node = names(lambda_t), mean = lambda_t) 

# tilted density
df_plot <- res$tilted_density[[1]]

ggplot2::ggplot() +
  ggplot2::geom_col(data = df_plot, ggplot2::aes(x = Z, y = Density, fill='Tilted density')) +
  ggplot2::geom_vline(data = means_df,
             ggplot2::aes(xintercept = mean, linetype = "Predicitive mean"),  linewidth = 1) +
    ggplot2::geom_vline(data = obs_df,
             ggplot2::aes(xintercept = Tot, linetype = "Observed data"),  linewidth = 1) +
  ggplot2::scale_colour_brewer(palette="Set1") +
  ggplot2::guides(fill='none')+
  ggplot2::labs(x = "Counts",
       y = "Tilted probability mass")

```

