---
title: "Simulated zero inflated Beta distribution example"
output: 
  rmarkdown::html_vignette:
       toc: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Simulated zero inflated Beta distribution example}
  %\usepackage[UTF-8]{inputenc}
---

## Overview

```{r, setup, message = FALSE, warning = FALSE}
set.seed(123)

library(gamlss)
library(gamlss.dist)
library(tidyverse)
library(boundHTS)

n_obs   <- 400
burn_in <- 150
time    <- seq_len(n_obs)
nodes <- c("A", "AA", "AB")

get_beta_params <- function(mean, prec) {
  
  var <- (mean * (1 - mean)) / (1 + prec)
  
  if(any(mean <= 0 | mean >= 1)) {
    stop("mean must be in (0,1)")
  }
  
  max_var <- mean * (1 - mean)
  if(any(var <= 0 | var >= max_var)) {
    stop("variance must be in (0, mean*(1-mean))")
  }
  
  phi <- (mean * (1 - mean)) / var - 1
  
  alpha <- phi * mean
  beta  <- phi * (1 - mean)
  
  data.frame(Alpha = alpha, Beta = beta)
}

```


This vignette demonstrates how to:

1. Simulate correlated zero inflated Beta (ZIB) distributed time series at the bottom level of a hierarchy
2. Aggregate these series to higher levels with additional variability
3. Estimate Bayesian state–space models at each level using **brms**
4. Perform rolling one-step-ahead forecasting efficiently by reusing compiled models

The example is intentionally simple (two bottom-level series and one aggregate),but the workflow generalizes to larger hierarchies.

### Hierarchical structure

We consider a two-level hierarchy:

- **Bottom level**: two correlated proportions (`AA`, `AB`)
- **Top level**: their aggregate (`A`) and the complement (`B = 1 - A`) with noise.

All series take values in (0, 1) and are modelled using ZIB distributions. Temporal dependence is introduced on the logit scale.


## Step 1: Generate zero inflated Beta distributed data 

We want to simulate data that is correlated at the bottom level, and then aggregate the bottom layers to give subsequent layers.

### 1.1 Bottom-level dynamics (correlated AR(1) processes)

We introduce temporal dependence on the **logit scale** to ensure that the resulting series remain in (0, 1) after transformation.

```{r, bottom-level-ar1}
# Bottom-level nodes: AA and AB
target_mean <- c(0.10, 0.05) # we want lower proportions for this simulation
mu <- qlogis(target_mean) # work backwards to estimate coefficients

beta1 <- c(0.05, 0.05)
beta0 <- mu * (1 - beta1) # ensures mean is stationary

# AR noise
sd_AA <- sd_AB <- 0.15
rho <- -0.6

Sigma <- matrix(
  c(sd_AA^2, rho * sd_AA * sd_AB,
    rho * sd_AA * sd_AB, sd_AB^2),
  nrow = 2
)

logit_mat <- matrix(NA_real_, n_obs, 2)
logit_mat[1, ] <- rnorm(2)

# Simulate correlated AR(1) dynamics on the logit scale
for (t in 2:n_obs) {
  logit_mat[t, ] <-
    mu +
    beta1 * (logit_mat[t - 1, ] - mu) +
    MASS::mvrnorm(1, mu = c(0, 0), Sigma = Sigma)
}

# Remove burn-in
logit_mat <- logit_mat[-seq_len(burn_in), ]

# Transform to proportions
AA <- plogis(logit_mat[, 1])
AB <- plogis(logit_mat[, 2])

# Add in zero inflation
zi_AA <- 0.05  # P(x=0) = 5%
zi_AB <- 0.05  # P(x=0) = 5%

inflate_with_zero <- function(mu, zi) {
  u <- runif(length(mu))
  y <- numeric(length(mu))

  # Structural zero
  y[u < zi] <- 0

  # Latent AR part
  idx <- u >= zi
  y[idx] <- mu[idx] 
  return(y)
}

AA_zero <- inflate_with_zero(
  mu  = AA,
  zi = zi_AA
)

AB_zero <-inflate_with_zero(
  mu  = AB,
  zi = zi_AB
)

plot(1:250, AA_zero, type='l')
lines(1:250, AA, col='red')
plot(1:250, AB_zero, type='l')

```

### 1.2 Dependence diagnostics

```{r, dependence-check}
# Dependence between bottom-level nodes
stats::cor.test(AA_zero, AB_zero, method = "kendall")
```

### 1.3 Aggregation to the top level
At the next level, we aggregate the bottom-level proportions by averaging them.
To allow for additional variability at this level, we generate observations from a Beta distribution centered on the aggregate mean.

```{r, aggregation-A}
# Aggregation at node A with additional variability
A_mean <- 0.5 * AA_zero + 0.5 * AB_zero

kappa_A <- 200 # very tight to reduce noise
A <- stats::rbeta(
  n      = length(A_mean),
  shape1 = kappa_A * A_mean,
  shape2 = kappa_A * (1 - A_mean)
)

## Check white noise added to aggregation (+/- 5%)
plot(
  A - A_mean,
  type = "l",
  ylab = expression(A[t] - E(A[t])),
  xlab = "Time"
)
abline(h = 0, lty = 2)

# Top-level split
B <- 1 - A

```

### 1.4 Final data set
The resulting data set contains all hierarchical levels and will be used as input for the Bayesian models in Step 2.

```{r, top-level}

# Final data set
sim_data <- dplyr::tibble(
  Time = 1:c(n_obs-burn_in),
  A  = A,
  B  = B,
  AA = AA_zero, 
  AB = AB_zero
)

```



## Step 2: Generating model estimates

We estimate separate state–space models for:

- the **top-level series** (`A`)
- the **bottom-level series** (`AA`, `AB`)

### 2.1 Run the models

```{r, echo=TRUE, results='hide'}
nsim <- 1000
n_testyears <- 250 # withhold one year of data as an example

# Create a lagged variable used for modelling
sim_data <- sim_data %>%
  arrange(Time) %>%  # or whatever your ordering variable is
  mutate(
    across(c(AA, AB, A), ~ lag(.x), .names = "{.col}_lag")
  ) %>%
  filter(if_all(ends_with("_lag"), ~ !is.na(.)))

# Split data into test and training
training_data <- sim_data %>%
  filter(Time < n_testyears)
test_data <- sim_data %>%
  filter(Time == n_testyears)

# Model using gamlss
zib_AA <- gamlss(AA ~ AA_lag, sigma.formula = ~1, nu.formula = ~AA_lag,
                 family = BEZI,
                 data = training_data)
zib_AB <- gamlss(AB ~ AB_lag, sigma.formula = ~1, nu.formula = ~AB_lag,
                 family = BEZI,
                 data = training_data)
zib_A <- gamlss(A ~ A_lag, sigma.formula = ~1, nu.formula = ~A_lag,
                 family = BEZI,
                 data = training_data)

# Get one-step ahead predictions
mu_pred_A <- predict(zib_A, newdata = test_data, what = "mu", type = "response")
mu_pred_AA <- predict(zib_AA, newdata = test_data, what = "mu", type = "response")
mu_pred_AB <- predict(zib_AB, newdata = test_data, what = "mu", type = "response")
  
# Get precision of one-step ahead forecast
prec_beta_A  <- predict(zib_A, newdata = test_data, what = "sigma", type = "response")
prec_beta_AA  <- predict(zib_AA, newdata = test_data, what = "sigma", type = "response")
prec_beta_AB  <- predict(zib_AB, newdata = test_data, what = "sigma", type = "response")
  
# Get zero-inflation probability of one-step ahead forecast
zi_beta_A  <- predict(zib_A, newdata = test_data, what = "nu", type = "response")
zi_beta_AA  <- predict(zib_AA, newdata = test_data, what = "nu", type = "response")
zi_beta_AB  <- predict(zib_AB, newdata = test_data, what = "nu", type = "response")

# Create results dataframes
fc_A <- tibble(Node = "A", Time = n_testyears,
               `Forecast precision`= prec_beta_A,
               `Forecast zero` = zi_beta_A,
               `Point Forecast` = mu_pred_A)
fc_AA <- tibble(Node = "AA", Time = n_testyears,
                `Forecast precision`= prec_beta_AA,
                `Forecast zero` = zi_beta_AA,
                `Point Forecast` = mu_pred_AA)
fc_AB <- tibble(Node = "AB", Time = n_testyears,
                `Forecast precision`= prec_beta_AB,
                `Forecast zero` = zi_beta_AB,
                `Point Forecast` = mu_pred_AB)
  
forecasts <- bind_rows(fc_A, fc_AA, fc_AB) # one step ahead

# Generate samples of predictions
sim_logit_A <- tibble(Node = "A", Time = n_testyears,
                      sim_pred = rBEZI(n = 100, mu = mu_pred_A, sigma = sqrt(prec_beta_A), nu = zi_beta_A)) 
sim_logit_AA <- tibble(Node = "AA", Time = n_testyears,
                       sim_pred = rBEZI(n = 100, mu = mu_pred_AA, sigma = sqrt(prec_beta_AA), nu = zi_beta_AA)) 
sim_logit_AB <- tibble(Node = "AB", Time = n_testyears,
                       sim_pred = rBEZI(n = 100, mu = mu_pred_AB, sigma = sqrt(prec_beta_AB), nu = zi_beta_AB)) 
  
forecasts_sims <- bind_rows(sim_logit_A, sim_logit_AA, sim_logit_AB)

 
```


## Step 3: Coherent hierarchical forecasting

We estimate a coherent top-series probability density function (PDF) using a convolution and weighted Beta distributions. As inputs into this process we use:
- the **top-level series** (`A`) estimates
- the **bottom-level series** (`AA`, `AB`) estimates
- A fine grid over which the density exists
- Bottom-series weights

### Step 3.1: Generate probability density function via convolution
We will generate PDFs for the withheld data points to validate our method. You can loop over all years of you wish by updating the code to do so.

```{r}
## Get Beta parameters
beta_params_df <- get_beta_params(forecasts$`Point Forecast`, forecasts$`Forecast precision`)
beta_params_df$Node <- forecasts$Node
  
alpha_list <- list(beta_params_df %>% filter(Node %in% c("AA", "AB")) %>% pull(Alpha),
                   beta_params_df %>% filter(Node %in% c("A")) %>% pull(Alpha))

beta_list <- list(beta_params_df %>% filter(Node %in% c("AA", "AB")) %>% pull(Beta),
                   beta_params_df %>% filter(Node %in% c("A")) %>% pull(Beta))

zi_list <- list(forecasts %>% filter(Node %in% c("AA", "AB")) %>% pull(`Forecast zero`),
                forecasts %>% filter(Node %in% c("A")) %>% pull(`Forecast zero`))


# Set up weights
weights_list <- list(c(0.75, 0.25), 1)

# Set up expected values 
mu_theory <- forecasts %>% 
  filter(Node %in% c("A")) %>% 
  arrange(Node) %>% 
  pull(`Point Forecast`)

# Run convolution with tilting 
res <- run_convolution_with_tilting(
  family = "ZIB",
  groups = list(Bottom = c("AA", "AB"), Top = "A"),
  point = TRUE,
  mu_theory = mu_theory,
  z_values = seq(0, 1, length.out = 1000),
  alpha_list = alpha_list,
  beta_list = beta_list,
  zi_list = zi_list,
  weights_list = weights_list,
  n_draws = 2000,
  n_sims = 100
)

```


## Step 4: Visualise the tilted convoluted PDF and compare against the predictive mean

The figure presents the predictive density formed by convolving the estimated bottom-level weighted Beta distributions. The resulting density is subsequently tilted so that its expectation matches the corresponding one-step-ahead predictive mean, ensuring coherence between the full predictive distribution and the point forecast.

For a single time point:
```{r, echo=FALSE, fig.width=9, fig.height=5, dpi=150}
t_point <- 250

obs_point <- sim_data %>% 
  filter(Time==t_point) %>%
  mutate(Method="Observed data")

forecasts_mean <- forecasts %>% 
  dplyr::filter(Time == t_point) %>%
  dplyr::select(Time, Node, `Point Forecast`) %>%
  tidyr::pivot_wider(names_from = Node, values_from = `Point Forecast`) %>%
  dplyr::mutate(Method = 'Predicted mean')

exp_plot_df <- res$tilted_density[[1]] %>% mutate(Method= "Tilted density")


ggplot2::ggplot() +
  ggplot2::geom_line(data = exp_plot_df,
                ggplot2::aes(x = Z, y = Density, colour=Method), linewidth = 1) +
  ggplot2::geom_vline(data = forecasts_mean,
                ggplot2::aes(xintercept = A, colour = Method), linewidth = 1) +
  ggplot2::geom_vline(data = obs_point,
                ggplot2::aes(xintercept = A, colour = Method), linewidth = 1) +
  ggplot2::labs(x = "Proportion", y = "Density") +
  ggplot2::theme_minimal(base_size = 14) +
  ggplot2::theme(legend.position = "bottom")


```



