---
title: "Simulated zero inflated Beta distribution example"
output: 
  rmarkdown::html_vignette:
       toc: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Simulated zero inflated Beta distribution example}
  %\usepackage[UTF-8]{inputenc}
---

## Overview

```{r, setup, message = FALSE, warning = FALSE}
set.seed(123)

library(gamlss)
library(gamlss.dist)
library(tidyverse)
library(boundHTS)

n_obs   <- 400
burn_in <- 150
time    <- seq_len(n_obs)
nodes <- c("A", "AA", "AB")

beta_params <- function(mean, prec) {
  alpha <- prec*mean
  beta  <-  prec*(1-mean)
  return(c(alpha, beta))
}

```


This vignette demonstrates how to:

1. Simulate correlated zero inflated Beta (ZIB) distributed time series at the bottom level of a hierarchy
2. Aggregate these series to higher levels with additional variability
3. Estimate Bayesian state–space models at each level using **brms**
4. Perform rolling one-step-ahead forecasting efficiently by reusing compiled models

The example is intentionally simple (two bottom-level series and one aggregate),but the workflow generalizes to larger hierarchies.

### Hierarchical structure

We consider a two-level hierarchy:

- **Bottom level**: two correlated proportions (`AA`, `AB`)
- **Top level**: their aggregate (`A`) and the complement (`B = 1 - A`) with noise.

All series take values in (0, 1) and are modelled using ZIB distributions. Temporal dependence is introduced on the logit scale.


## Step 1: Generate zero inflated Beta distributed data 

We want to simulate data that is correlated at the bottom level, and then aggregate the bottom layers to give subsequent layers.

### 1.1 Bottom-level dynamics (correlated AR(1) processes)

We introduce temporal dependence on the **logit scale** to ensure that the resulting series remain in (0, 1) after transformation.

```{r, bottom-level-ar1}
# Bottom-level nodes: AA and AB
target_mean <- c(0.10, 0.05) # we want lower proportions for this simulation
mu <- qlogis(target_mean) # work backwards to estimate coefficients

beta1 <- c(0.05, 0.05)
beta0 <- mu * (1 - beta1) # ensures mean is stationary

# AR noise
sd_AA <- sd_AB <- 0.15
rho <- -0.6

Sigma <- matrix(
  c(sd_AA^2, rho * sd_AA * sd_AB,
    rho * sd_AA * sd_AB, sd_AB^2),
  nrow = 2
)

logit_mat <- matrix(NA_real_, n_obs, 2)
logit_mat[1, ] <- rnorm(2)

# Simulate correlated AR(1) dynamics on the logit scale
for (t in 2:n_obs) {
  logit_mat[t, ] <-
    mu +
    beta1 * (logit_mat[t - 1, ] - mu) +
    MASS::mvrnorm(1, mu = c(0, 0), Sigma = Sigma)
}

# Remove burn-in
logit_mat <- logit_mat[-seq_len(burn_in), ]

# Transform to proportions
AA <- plogis(logit_mat[, 1])
AB <- plogis(logit_mat[, 2])

# Add in zero inflation
zi_AA <- 0.05  # P(x=0) = 5%
zi_AB <- 0.05  # P(x=0) = 5%

inflate_with_zero <- function(mu, zi) {
  u <- runif(length(mu))
  y <- numeric(length(mu))

  # Structural zero
  y[u < zi] <- 0

  # Latent AR part
  idx <- u >= zi
  y[idx] <- mu[idx] 
  return(y)
}

AA_zero <- inflate_with_zero(
  mu  = AA,
  zi = zi_AA
)

AB_zero <-inflate_with_zero(
  mu  = AB,
  zi = zi_AB
)

plot(1:250, AA_zero, type='l')
lines(1:250, AA, col='red')
plot(1:250, AB_zero, type='l')

```

### 1.2 Dependence diagnostics

```{r, dependence-check}
# Dependence between bottom-level nodes
stats::cor.test(AA_zero, AB_zero, method = "kendall")
```

### 1.3 Aggregation to the top level
At the next level, we aggregate the bottom-level proportions by averaging them.
To allow for additional variability at this level, we generate observations from a Beta distribution centered on the aggregate mean.

```{r, aggregation-A}
# Aggregation at node A with additional variability
A_mean <- 0.5 * AA_zero + 0.5 * AB_zero

kappa_A <- 200 # very tight to reduce noise
A <- stats::rbeta(
  n      = length(A_mean),
  shape1 = kappa_A * A_mean,
  shape2 = kappa_A * (1 - A_mean)
)

## Check white noise added to aggregation (+/- 5%)
plot(
  A - A_mean,
  type = "l",
  ylab = expression(A[t] - E(A[t])),
  xlab = "Time"
)
abline(h = 0, lty = 2)

# Top-level split
B <- 1 - A

```

### 1.4 Final data set
The resulting data set contains all hierarchical levels and will be used as input for the Bayesian models in Step 2.

```{r, top-level}

# Final data set
sim_data <- dplyr::tibble(
  Time = 1:c(n_obs-burn_in),
  A  = A,
  B  = B,
  AA = AA_zero, 
  AB = AB_zero
)

```



## Step 2: Generating model estimates

We estimate separate state–space models for:

- the **top-level series** (`A`)
- the **bottom-level series** (`AA`, `AB`)

### 2.1 Run the models

```{r, echo=TRUE, results='hide'}
forecasts_sims <- forecasts <- data.frame()
nsim <- 1000
n_testyears <- tail(sim_data$Time)

# Create a lagged variable used for modelling
sim_data$AA_lag <- dplyr::lag(sim_data$AA)
sim_data$AB_lag <- dplyr::lag(sim_data$AB)
sim_data$A_lag <- dplyr::lag(sim_data$A)


# Model using gamlss
for(t in seq_along(n_testyears)) {
  
  # Calculate training data
  train_idx <- which(sim_data$Time < n_testyears[t])
  training_data <- sim_data[train_idx[-1], , drop = FALSE] # drop NA lag
  
  # Fit models
  zib_AA <- gamlss(AA ~ AA_lag, 
                   sigma.formula = ~1,
                   nu.formula = ~AA_lag,
                   family = BEZI,
                   data = training_data)
  zib_AB <- gamlss(AB ~ AB_lag, 
                   sigma.formula = ~1,
                   nu.formula = ~AB_lag,
                   family = BEZI,
                   data = training_data)
  zib_A <- gamlss(A ~ A_lag, 
                  sigma.formula = ~1,
                  nu.formula = ~A_lag,
                  family = BEZI,
                  data = training_data)

  if(t==1) {
    pred_A <- tibble(`Point Forecast` = predict(zib_AA, type = "response")) %>% 
      mutate(Node = "A", Time = 2:length(train_idx))
    pred_AA <- tibble(`Point Forecast` =  predict(zib_AA, type = "response")) %>% 
      mutate(Node = "AA", Time = 2:length(train_idx))
    pred_AB <- tibble(`Point Forecast` =  predict(zib_AA, type = "response")) %>% 
      mutate(Node = "AB", Time = 2:length(train_idx))
    
    forecasts <- bind_rows(forecasts, pred_A, pred_AA, pred_AB) # training data estimates
    
  }
  
  # Get one-step ahead predictions
  mu_pred_A <- predict(zib_A, newdata = sim_data[n_testyears[t], ],
                       what = "mu", type = "response")
  mu_pred_AA <- predict(zib_AA, newdata = sim_data[n_testyears[t], ],
                        what = "mu", type = "response")
  mu_pred_AB <- predict(zib_AB, newdata = sim_data[n_testyears[t], ],
                        what = "mu", type = "response")
  
  # Get precision of one-step ahead forecast
  prec_beta_A  <- predict(zib_A, newdata = sim_data[n_testyears[t], ],
                        what = "sigma", type = "response")
  prec_beta_AA  <- predict(zib_AA, newdata = sim_data[n_testyears[t], ],
                        what = "sigma", type = "response")
  prec_beta_AB  <- predict(zib_AB, newdata = sim_data[n_testyears[t], ],
                        what = "sigma", type = "response")
  
  # Get zero-inflation probability of one-step ahead forecast
  zi_beta_A  <- predict(zib_A, newdata = sim_data[n_testyears[t], ],
                        what = "nu", type = "response")
  zi_beta_AA  <- predict(zib_AA, newdata = sim_data[n_testyears[t], ],
                        what = "nu", type = "response")
  zi_beta_AB  <- predict(zib_AB, newdata = sim_data[n_testyears[t], ],
                        what = "nu", type = "response")
  
  # Generate samples of predictions
  sim_pred_A <- rBEZI(n = 100, mu = mu_pred_A,
                      sigma = sqrt(prec_beta_A), nu = zi_beta_A)
  sim_pred_AA <- rBEZI(n = 100, mu = mu_pred_AA,
                      sigma = sqrt(prec_beta_AA), nu = zi_beta_AA)
  sim_pred_AB <- rBEZI(n = 100, mu = mu_pred_AB,
                      sigma = sqrt(prec_beta_AB), nu = zi_beta_AB)
  
  # Create results dataframes
  fc_A <- tibble(Node = "A",
                 Time = n_testyears[t],
                 `Forecast precision`= prec_beta_A,
                 `Forecast zero` = zi_beta_A,
                 `Point Forecast` = mu_pred_A)
  fc_AA <- tibble(Node = "AA",
                 Time = n_testyears[t],
                 `Forecast precision`= prec_beta_AA,
                 `Forecast zero` = zi_beta_AA,
                 `Point Forecast` = mu_pred_AA)
  fc_AB <- tibble(Node = "AB",
                 Time = n_testyears[t],
                 `Forecast precision`= prec_beta_AB,
                 `Forecast zero` = zi_beta_AB,
                 `Point Forecast` = mu_pred_AB)
  
  forecasts <- bind_rows(forecasts, fc_A, fc_AA, fc_AB)
 
  # Create results dataframes
  sim_logit_A <- tibble(Node = "A",
                        Time = n_testyears[t],
                        sim_pred = sim_pred_A) 
  sim_logit_AA <- tibble(Node = "AA",
                        Time = n_testyears[t],
                        sim_pred = sim_pred_AA) 
  sim_logit_AB <- tibble(Node = "AB",
                        Time = n_testyears[t],
                        sim_pred = sim_pred_AB) 
  
  forecasts_sims <- bind_rows(forecasts_sims, sim_logit_A, sim_logit_AA, sim_logit_AB)
  
}

```


## Step 3: Coherent hierarchical forecasting

We estimate a coherent top-series probability density function (PDF) using a convolution and weighted Beta distributions. As inputs into this process we use:
- the **top-level series** (`A`) estimates
- the **bottom-level series** (`AA`, `AB`) estimates
- A fine grid over which the density exists
- Bottom-series weights

### Step 3.1: Generate probability density function via convolution
We will generate PDFs for the withheld data points to validate our method. You can loop over all years of you wish by updating the code to do so.

```{r}
z_values <- seq(0, 1, length.out = 1000) # density grid
top_node <- "A"
bottom_nodes <- c("AA", "AB")
weights_bottom <- c(0.75, 0.25)
  
# Create results df
n_sims <- 100
n_draws <- 100
n_bottom <- 2

rec_list <- lapply(seq_along(n_testyears), function(i) {

  rec_local <- tibble()   # initialize

  t_i <- n_testyears[i]

  ## Get predictions
  P_top <- forecasts %>% 
    filter(Time == t_i, Node == top_node) %>% 
    pull(`Point Forecast`)

  P_bottom <- forecasts %>% 
    filter(Time == t_i, Node %in% bottom_nodes) %>% 
    pull(`Point Forecast`)

  ## Get precision
  prec_top <- forecasts %>% 
    filter(Time == t_i, Node == top_node) %>% 
    pull(`Forecast precision`)

  prec_bottom <- forecasts %>% 
    filter(Time == t_i, Node %in% bottom_nodes) %>% 
    pull(`Forecast precision`)

    ## Get precision
  zoi_top <- forecasts %>% 
    filter(Time == t_i, Node == top_node) %>% 
    pull(`Forecast zero`)
  
  zoi_bottom <- forecasts %>% 
    filter(Time == t_i, Node %in% bottom_nodes) %>% 
    pull(`Forecast zero`)
  
  ## Get Beta parameters
  top_params <- beta_params(P_top, prec_top)

  bottom_params <- list(
    beta_params(P_bottom[1], prec_bottom[1]),
    beta_params(P_bottom[2], prec_bottom[2])
  )

  weighted_samps <- array(NA, dim = c(n_sims, n_draws, n_bottom))

  for (b in 1:n_bottom) {

    weighted_samps[, , b] <- matrix(
      rZIB_4p(
        n_mc = n_sims * n_draws,
        alpha_point = bottom_params[[b]][1],
        beta_point = bottom_params[[b]][2],
        zoi_point = zoi_top,
        0,
        weights_bottom[b]
      ),
      nrow = n_sims,
      ncol = n_draws
    )

    dens <- dbeta(
      z_values,
      bottom_params[[b]][1],
      bottom_params[[b]][2]
    )

    bottom_tmp <- tibble(
      Node = bottom_nodes[b],
      Time = t_i,
      Z = z_values,
      Density = dens
    )

    rec_local <- rbind(rec_local, bottom_tmp)
  }

  ## Convolution
  Density_top <- ZIB_convolution(
    z_values = z_values,
    alpha_input = bottom_params[[n_bottom]][1],
    beta_input = bottom_params[[n_bottom]][2],
    zi_input = zoi_bottom[n_bottom],
    weighted_samps = weighted_samps,
    weights = weights_bottom[n_bottom],
    point=TRUE
  )

  tmp <- tibble(
    Node = top_node,
    Time = t_i,
    Z = z_values,
    Density = Density_top
  )

  rec_local <- rbind(rec_local, tmp)

  rec_local  # return
})

rec_df <- dplyr::bind_rows(rec_list)


```

### Step 3.2: Tilt the convoluted PDF towards the predictive mean

```{r}
# Define node set for A-group
all_nodes <- c("A", "AA", "AB")
tilted_nodes <- c("A")

# Storage
nu_exp <- exp_samps <- f_tilde_exp <- list()

# Loop over time
for (t in seq_along(n_testyears)) {
  
  t_i <- n_testyears[t]

  # Extract convolution densities for these nodes
  rec_dens_df <- rec_df %>%
    dplyr::filter(Time == t_i) %>%
    tidyr::pivot_wider(names_from = Node, values_from = Density) %>%
    dplyr::select(Z, dplyr::all_of(tilted_nodes))

  # Get predictions
  mu_theory <- forecasts %>% 
    filter(Time == t_i, Node %in% c(top_node, bottom_nodes)) %>% 
    pull(`Point Forecast`)

  # y grid
  y_vals <- sort(unique(rec_dens_df$Z))

  # Initialize results
  nu_star_vec <- numeric(length(tilted_nodes))
  tilted_samps <- matrix(NA, nrow = 5000, ncol = length(tilted_nodes))
  f_tilted <- matrix(NA, nrow = length(z_values), ncol = length(tilted_nodes))
  colnames(tilted_samps) <- colnames(f_tilted) <- tilted_nodes

  for (i in seq_along(tilted_nodes)) {
    name <- tilted_nodes[i]

    # Extract and normalise convolution base density
    f_y <- as.numeric(rec_dens_df[[name]])
    f_y <- f_y / pracma::trapz(y_vals, f_y)

    # Tilt density
    tilted_dens <- tilt_density(mu_theory[i], y_vals, f_y, discrete=FALSE) # tilt density
  
    nu_star_vec[i] <- tilted_dens$nu_star

    # Tilted density + samples
    f_tilted[, i] <- tilted_dens$f_tilted

    tilted_samps[, i] <- tilted_dens$tilted_samps
    }

  # Save per-t results
  nu_exp[[t]]      <- nu_star_vec
  exp_samps[[t]]   <- tilted_samps
  f_tilde_exp[[t]] <- f_tilted
}


```

## Step 4: Visualise the tilted convoluted PDF and compare against the predictive mean

The figure presents the predictive density formed by convolving the estimated bottom-level weighted Beta distributions. The resulting density is subsequently tilted so that its expectation matches the corresponding one-step-ahead predictive mean, ensuring coherence between the full predictive distribution and the point forecast.

For a single time point:
```{r, echo=FALSE, fig.width=9, fig.height=5, dpi=150}
t_point <- 250

forecasts_mean <- forecasts %>% 
  dplyr::filter(Time == t_point) %>%
  dplyr::select(Time, Node, `Point Forecast`) %>%
  tidyr::pivot_wider(names_from = Node, values_from = `Point Forecast`) %>%
  dplyr::mutate(Method = 'Predicted mean')

conv_dens <- rec_df %>%
  dplyr::filter(Time == t_point & Node==tilted_nodes) %>%
  tidyr::pivot_wider(names_from = Node, values_from = Density) %>%
  dplyr::mutate(Method = 'Convolution')

exp_plot_df <- dplyr::bind_rows(lapply(seq_along(f_tilde_exp), function(t) {
  tibble::tibble(
    x = z_values,
    Density = as.vector(f_tilde_exp[[t]]),
    Method = "Exp. tilted density",
    Time = n_testyears[t])})) %>%
  dplyr::mutate(Density = ifelse(is.na(Density)==TRUE, 0, Density)) %>%
  dplyr::filter(Time==t_point)


ggplot2::ggplot() +
  ggplot2::geom_line(data = exp_plot_df,
                ggplot2::aes(x = x, y = Density, colour = Method), linewidth = 1) +
  ggplot2::geom_line(data = conv_dens,
                ggplot2::aes(x = Z, y = A, colour = Method), linewidth = 1) +
  ggplot2::geom_vline(data = forecasts_mean,
                ggplot2::aes(xintercept = A, colour = Method), linewidth = 1) +
  ggplot2::scale_colour_manual( values = c(
    "Exp. tilted density" = "#1b9e77",
    "Convolution" = "#d95f02",
    "Predicted mean" = "#7570b3"))+
  ggplot2::labs(x = "Proportion", y = "Density") +
  ggplot2::theme_minimal(base_size = 14)

```



