---
title: "Simulated Beta distribution example"
output: 
  rmarkdown::html_vignette:
       toc: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Simulated Beta distribution example}
  %\usepackage[UTF-8]{inputenc}
---

## Install package
```{r, install, message = FALSE, warning = FALSE, eval=FALSE}
install.packages("remotes")
remotes::install_github("hannahcomiskey/boundHTS")

```


## Overview

This vignette demonstrates how to:

1. Simulate correlated Beta-distributed time series at the bottom level of a hierarchy
2. Aggregate these series to higher levels with additional variability
3. Estimate Bayesian state–space models at each level using **JAGS**
4. Perform rolling one-step-ahead forecasting efficiently by reusing compiled models

The example is intentionally simple (two bottom-level series and one aggregate), but the workflow generalizes to larger hierarchies.

```{r, setup, message = FALSE, warning = FALSE}

library(tidyverse)
library(dplyr)
library(tidyr)
library(boundHTS)
library(forecast)
library(ggplot2)

set.seed(123)
```

### Some helper functions for later
```{r helper functions}
inv_logit <- function(x) {
 p <- 1 / (1 + exp(-x))
 return(p)
}

inv_logit_delta <- function(mu, var) {
  prop_var <- (mu * (1 - mu))^2 * var
  return(prop_var)
} 

beta_params <- function(mean, var) {
  alpha <- (((mean*(1-mean))/var)-1) * mean
  beta  <- (((mean*(1-mean))/var)-1) * (1-mean)
  return(c(alpha, beta))
}

# Function to fit models and forecast h-step ahead
fit_and_forecast <- function(x, node_name, h = 1, nsim = 5000) {
  
  mod <- auto.arima(x)
  
  fc <- forecast(mod, h = h) %>% as_tibble()
  
  sim_vals <- replicate(nsim, simulate(mod, nsim = h, future = TRUE))
  var_beta <- var(sim_vals)
  
  fc <- fc %>%
    mutate(Node = node_name,
           Time = n_testyears,
           logit_var = var_beta) %>%
    rename(logit_mean = `Point Forecast`)
  
  return(fc)
}

```


### Hierarchical structure

We consider a two-level hierarchy:

- **Bottom level**: two correlated proportions (`AA`, `AB`)
- **Top level**: their aggregate (`A`) and the complement (`B = 1 - A`) with noise.

All series take values in (0, 1) and are modelled using Beta distributions. Temporal dependence is introduced on the logit scale.


## Step 1: Load simulated Beta distributed data 

```{r}
head(beta_sim_data)

```


## Step 2: Generating model estimates

We estimate separate autoregressive models for:

- the **top-level series** (`A`)
- the **bottom-level series** (`AA`, `AB`)

We will use the logit-transformed proportions and an ARMA model to capture the temporal changes

### 2.1 Calculate logit of data

```{r}
logit_sim_data <- beta_sim_data %>% 
  mutate(logit_A = log(A/(1-A)),
         logit_AA = log(AA/(1-AA)),
         logit_AB = log(AB/(1-AB)))

```

### 2.2 Run the models

```{r}
nsim <- 1000
n_testyears <- 250 # withhold one year of data as an example

# Split data into test and training
train_idx <- which(logit_sim_data$Time < n_testyears)
training_data <- logit_sim_data[train_idx, , drop = FALSE]

# Fit demo models
fc_A <- fit_and_forecast(training_data$logit_A, node_name = "A")
fc_AA <- fit_and_forecast(training_data$logit_AA, node_name = "AA")
fc_AB <- fit_and_forecast(training_data$logit_AB, node_name = "AB")

forecasts <- bind_rows(fc_A, fc_AA, fc_AB)

# Back transform logit-transformed quantities to proportions
prop_var <- inv_logit_delta(inv_logit(forecasts$logit_mean), forecasts$logit_var)

forecasts <- forecasts %>%
  mutate(proportion = inv_logit(forecasts$logit_mean),
         prop_var = prop_var) %>%
  select(Time, Node, logit_mean, logit_var, proportion, prop_var)

head(forecasts)

```


## Step 3: Coherent hierarchical forecasting

We estimate a coherent top-series probability density function (PDF) using a convolution and weighted Beta distributions. As inputs into this process we use:
- the **top-level series** (`A`) estimates
- the **bottom-level series** (`AA`, `AB`) estimates
- A fine grid over which the density exists
- Bottom-series weights

### Step 3.1: Generate probability density function via convolution
We will generate PDFs for the withheld data points to validate our method. You can loop over all years of you wish by updating the code to do so.

```{r}
z_values <- seq(0, 1, length.out = 1000) # density grid
top_node <- "A"
bottom_nodes <- c("AA", "AB")
weights_bottom <- c(0.75, 0.25)
  
# Create results df
n_sims <- 100
n_draws <- 100
n_bottom <- 2

## Get Beta parameters
top_params <- beta_params(mean = as.numeric(forecasts[which(forecasts$Node=="A"), "proportion"]),
                          var = as.numeric(forecasts[which(forecasts$Node=="A"), "prop_var"]))
AA_params <- beta_params(mean = as.numeric(forecasts[which(forecasts$Node=="AA"), "proportion"]),
                         var = as.numeric(forecasts[which(forecasts$Node=="AA"), "prop_var"]))
AB_params <- beta_params(mean = as.numeric(forecasts[which(forecasts$Node=="AB"), "proportion"]),
                         var = as.numeric(forecasts[which(forecasts$Node=="AB"), "prop_var"]))

bottom_params <- list(AA_params, AB_params)

# empty array to store results
weighted_samps <- array(NA, dim = c(n_sims, n_draws, n_bottom))

# Rescale bottom series onto the [0,w] interval for aggregation later
for (b in 1:n_bottom) {
  weighted_samps[, , b] <- as.matrix(ExtDist::rBeta_ab(n = n_sims * n_draws,
                                                    shape1 = bottom_params[[b]][1],
                                                    shape2 = bottom_params[[b]][2],
                                                    a = 0,
                                                    b = weights_bottom[b]),
                                     nrow = n_sims,
                                     ncol = n_draws)
}

## Convolution of bottom series
Density_top <- Beta_convolution(z_values = z_values, # grid of values to estimate density over
                                alpha_input =  c(AA_params[1], AB_params[1]), # Parameters of bottom series
                                beta_input  = c(AA_params[2], AB_params[2]),
                                weighted_samps = weighted_samps, # Rescaled bottom series
                                weights = weights_bottom, # weights
                                point=TRUE)
  
# store density results
top_dens_df <- tibble(Node = top_node, 
                      Time = n_testyears,
                      Z = z_values,
                      Density = Density_top)

```

### Step 3.2: Tilt the convoluted PDF towards the predictive mean

```{r}
# Define node set for A-group
all_nodes <- c("A", "AA", "AB")
tilted_nodes <- c("A")

# Extract convolution densities for these nodes
rec_dens_df <- top_dens_df %>%
  filter(Time == n_testyears) %>%
  pivot_wider(names_from = Node, values_from = Density)

# Get predictions
mu_theory <- forecasts %>% 
  filter(Time == n_testyears) %>% 
  pull(proportion)

# y grid
y_vals <- sort(unique(rec_dens_df$Z))

# Initialize results
nu_star_vec <- numeric(length(tilted_nodes))
tilted_samps <- matrix(NA, nrow = 5000, ncol = length(tilted_nodes))
f_tilted <- matrix(NA, nrow = length(z_values), ncol = length(tilted_nodes))
colnames(tilted_samps) <- colnames(f_tilted) <- tilted_nodes

# Extract and normalise convolution base density
f_y <- as.numeric(rec_dens_df[["A"]])
f_y <- f_y / pracma::trapz(y_vals, f_y)

# Tilt density
tilted_dens <- tilt_density(mu_theory[1], y_vals, f_y, discrete=FALSE) # tilt density
nu_star_vec <- tilted_dens$nu_star

# Tilted density + samples
f_tilted <- tilted_dens$f_tilted
tilted_samps <- tilted_dens$tilted_samps


```

## Step 4: Visualise the tilted convoluted PDF and compare against the predictive mean

The figure presents the predictive density formed by convolving the estimated bottom-level weighted Beta distributions. The resulting density is subsequently tilted so that its expectation matches the corresponding one-step-ahead predictive mean, ensuring coherence between the full predictive distribution and the point forecast.

For a single time point:
```{r, echo=FALSE, fig.width=9, fig.height=5, dpi=150}
t_point <- 250

obs_point <- beta_sim_data %>% 
  filter(Time==t_point) %>%
  mutate(Method="Observed data")

forecasts_mean <- forecasts %>% 
  filter(Time == t_point) %>%
  select(Time, Node, proportion) %>%
  pivot_wider(names_from = Node, values_from = proportion) %>%
  mutate(Method = 'Predicted mean')

conv_dens <- top_dens_df %>%
  pivot_wider(names_from = Node, values_from = Density) %>%
  mutate(Method = 'Convolution')

exp_plot_df <- tibble::tibble(x = z_values,
                              Density = as.vector(f_tilted),
                              Method = "Exp. tilted density",
                              Time = n_testyears) %>%
  mutate(Density = ifelse(is.na(Density)==TRUE, 0, Density))


ggplot() +
  geom_line(data = exp_plot_df,
                aes(x = x, y = Density, colour = Method), linewidth = 1) +
  geom_line(data = conv_dens,
                aes(x = Z, y = A, colour = Method), linewidth = 1) +
  geom_vline(data = forecasts_mean,
                aes(xintercept = A, colour = Method), linewidth = 1) +
  geom_vline(data = obs_point,
                aes(xintercept = A, colour = Method), linewidth = 1) +
  scale_colour_manual( values = c("Exp. tilted density" = "#1b9e77",
                                           "Convolution" = "#d95f02",
                                           "Bottom density" = "#626f02",
                                           "Predicted mean" = "#7570b3",
                                           "Observed data" = "#1950b3")) +
  labs(x = "Proportion", y = "Density") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom")

```




